Financialization of Commodities: an Asset Pricing Perspective

### 2024-06-11 others/paper/custom/main.tex
Track 'others' folder

\documentclass[12pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[letterpaper, left = 1in, right = 1in, top = 1.5in, bottom = 1.5in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Financialization of Commodities: an Asset Pricing Perspective},
            pdfkeywords={Commodity futures; financialization; hedging pressure; financial crisis},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{natbib}
\bibliographystyle{jf}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

  \title{Financialization of Commodities: an Asset Pricing Perspective}
    \author{}
    \date{}

@@

@@

The last twenty years have seen major upheavals in the commodity markets
with the onset of financialization\footnote{Starting around 2004, with a
  view of commodity markets as an emerging asset class and in an effort
  to hedge against inflation and seek diversification benefits
  \citep{buyuksahin_speculators_2014, singleton_investor_2013},
  institutional investors sent forth an unprecedented flow of capital
  into commodity futures markets. This phenomenon is commonly referred
  to as the financialization of commodities
  \citep{domanski_financial_2007}.} as well as the period of the 2008
financial crisis and its aftermath. Financialization appears to have
affected a number of economic sectors, including agriculture and energy,
and its impact has been hotly debated in the policy, legislative,
regulatory, and academic spheres. A vigorous legislative debate has led
to regulatory changes accompanied by an equally intense policy and
academic debate across a number of disciplines about the perceived
effects on commodity prices of the unprecedented inflow of institutional
funds brought into commodity futures markets by new index type
investors\footnote{See section \ref{background} for a detailled
  discussion.}. In response to immediate policy concerns the academic
debate was initially framed rather narrowly. In this paper we take a
fundamentally different empirical approach that is broader in scope. We
consider the entire cross-section of actively traded commodities on
futures markets and use a futures based asset pricing framework that
includes factors constructed using both price and non-price attributes.

Commodity futures markets have a history of being the classic physical
commodity risks shifting conduit for commodity specialists, and
financialization may have altered this long standing paradigm with the
entry of long only commodity index traders (CITs) with very different
hedging and volume demands from traditional participants. This change
could have altered the hedging ontology in these markets and our asset
pricing framework is well suited to detect these more subtle effects
driven by changes in the nature of traders. Factor models are well
suited to isolating common driving factors and detecting co-movement and
building on theory we develop a bespoke commodity-based asset pricing
framework that seems well adapted to study these two issues
simultaneously.

@@

We study the performance of these models on a set of twenty-four US
traded commodities and six UK traded (LME exchange) base metals over the
1997/2017 period that we divide into four distinct sub-periods. The
first, the pre-financialization phase, spans the 1997/2003 period; the
second, the financialization phase, spans the 2004/2007 period\footnote{Starting
  point based on earlier studies
  \citep{baker_financialization_2014, christoffersen_factor_2018}.}; the
third, 2008/2013, encompasses both the financial crisis and the loose
monetary policy regimes that followed; and the fourth spans the period
starting in 2014 to present.\\
Consistent with earlier studies, we observe that mean returns for the US
traded commodities are sharply higher over the financialization period
compared to pre-financialization and this pattern extends to UK traded
metals.

The arrival of financial investors raises the question of whether they
functioned as liquidity providers or liquidity demanders. Arguments have
been made for both sides with \citet{moskowitz_time_2012} arguing that
financial investors provide liquidity while
\citet{Kangtaletwopremiums2017} argue that they demand liquidity with
\citet{cheng_convective_2014} showing that they in fact transferred risk
to hedgers during the 2008 financial crisis. We study this issue by
extending the Keynesian individual hedging pressure paradigm by
incorporating aggregate CHP (\textbf{CHP}), a market wide measure of
hedging pressure first considered in \citet{hong_what_2012}. Our use of
this measure is motivated by the cross sectional nature of
financialization
\citep{basak_model_2016, cheng_convective_2014, le_futures_2017}. This
paradigm implies that returns for individual commodities should be high
in periods where \textbf{CHP} is low (\textbf{backwardation}) and low in
periods where \textbf{CHP} is high (\textbf{contango}). Over the first
period the cross sectional average returns during \textbf{backwardation}
for both the US and UK commodities are positive with those for the US
statistically significant, and significantly higher than during
\textbf{contango}, providing broad support for an aggregate version of
the Keynesian hedging pressure hypothesis. Over the second period this
pattern reverses, with both the US and UK commodities having significant
positive returns during \textbf{contango} with these returns being
significantly higher than during \textbf{backwardation}. This change in
pattern suggests that the onset of financialization altered the
traditional risk shifting nature of hedger's behaviour implied by the
Keynesian hypothesis and suggests that financial investors may have been
liquidity demanders during financialization leading hedgers to become
liquidity providers.

\citet{goldstein_commodity_2017} as well as
\citet{goldstein_speculation_2014} argue that the arrival of financial
investors has an impact on the extent of normal backwardation in
commodity futures markets. Based on their analysis we would expect our
average CHP factor, which is based on backwardation measured by hedger's
positions, as well our term structure factor, which is based on
backwardation measured by roll yield, to be best suited to measure the
impact of financialization.\\
We analyze the pricing dynamics over the different periods by first
studying the time series pricing performance of the various factors on
all the US traded commodities with the factors constructed using the
latter set of commodities. The market based factor outperforms the other
three in the first period while the CHP factor shows the greatest
improvement in performance in the second period with a four-fold
increase in pricing performance. The performance of the market factor is
driven by agricultural commodities and metals in both the first and
second period and the dramatic improvement of the CHP factor in the
second period is driven by precious and precious-base metals. The best
performing commodities for the market factor exhibited approximately the
same level of price co-movement both before and during financialization.
In contrast, the best performing commodities for the CHP factor
exhibited very low pair-wise correlation in the first period, around the
average for US traded commodities, while during financialization the
average pairwise correlation between the top performing CHP factor
commodities is higher than that for the market factor and double of the
average for all the US traded commodities. This increase is due to the
dramatic rise in average pairwise correlation amongst the US traded
metals all of which exhibited high R\textsuperscript{2} with respect to
the CHP factor during financialization which in turn is due to their
common low level of CHP over this period. Thus the CHP factor model is
able to establish the strengthening link between a non-price attribute
and price dynamics in the US metal sub-sector during financialization.\\
We explore this issue further by constructing the four factors with the
top eight commodities for a given factor (factor picks) in each
period\footnote{See section \ref{methods} for further details.} and
evaluating their performance on these selected commodities as well as on
all US traded commodities. The results of this exercise strongly suggest
that the hedging pressure co-movement and price co-movement link was
concentrated in the metals sub-sector.

The UK traded base metals provide a useful set of assets to test how
widespread were the changes in pricing dynamics brought about by the
onset of financialization. To that end, we examine the performance of
the four factors on the six UK metals, when constructed from each set of
factor picks. The results show that there was co-movement in global
metals returns in the second period and that this co-movement could be
detected by our hedging pressure based model.

The onset of the financial crisis and the monetary policy regimes that
followed, on the other hand, also appear to have induced significant
changes in pricing dynamics. Returns on both US and UK traded
commodities fell dramatically over both the third and fourth periods
with the pattern of return for the US traded commodities across
\textbf{CHP} regimes reverting to the Keynesian paradigm. The pricing
results indicate the emergence of a systematic factor across the entire
cross-section of the US commodities as well as strong evidence of
cross-market linkages.

Commodity financialization raised societal concerns, leading to
legislative action intended to alter the nature and functioning of these
markets. The debate was initially framed around adequacy of speculation,
the burning issue of the day, and the consequent analysis focused on the
more mechanical effects of financialization. With the benefit of
hindsight, it appears this approach was perhaps too narrow and it now
seems necessary to address financialization from a broader perspective.
Commodity price dynamics appear to have altered substantially in quite
different ways over the financialization and the financial crisis
periods and our commodity futures based asset pricing approach seems
able to provide new insights into the nature of these changes:
financialization was a phenomenon endogenous to the commodity markets
transmitted via the commodity futures markets and mainly concentrated in
a specific sector (metals) while the crisis and its aftermath seems to
have delivered an exogenous shock across the entire cross-section of
global commodities. We provide an empirical complement to a new stream
of theoretical studies that try to model the impact of financialization
on various aspects of commodity futures markets\footnote{\citet{etula_broker-dealer_2013},
  \citet{acharya_limits_2013}, \citet{cheng_convective_2014},
  \citet{leclercq_equilibrium_2014}, \citet{sockin_informational_2015},
  \citet{goldstein_speculation_2014},
  \citet{EkelandSpeculationcommodityfutures2016},
  \citet{goldstein_commodity_2017} as well as \citet{le_futures_2017},
  one of the few empirical studies on the impact of financialization on
  commodity-price cross-moments.} and demonstrate that the effects of
financialization extend beyond the mechanical effects induced by
indexation as outlined in \citet{basak_model_2016} as well as
\citet{tang_index_2012}.

@@

The period from 1998 to 2008 saw an influx of investment into commodity
index linked products most of which made its way into the commodity
futures market. Exposure to commodity indexes whose goal was to track
the broad movement of commodity prices could be accessed by means of
exchange traded funds and exchange traded notes and at least \$100
billion of new investment flowed into these products over the 1998-2008
period with this financialization phenomenon leading to substantial
legislative changes. Concerns over the consequences of financialization
were behind Rule 76 FR 4752 issued by the US Commodity Futures Trading
Commission (CFTC) on January 26\textsuperscript{th}, 2011. This
provision emanates from the Dodd-Frank Wall Street and Consumer
Protection Act of 2010 (Title VII, Section 737) that mandates the CFTC
to use position limits to restrict the flow of speculative capital into
a number of commodity markets. The Rule was approved in a close 3-2 vote
and the ensuing rulemaking process was extremely contentious with
several commissioners (Michael Dunn and Scott D. O'Malia in particular)
expressing reservations about the lack of supporting evidence and the
Rule also triggering thousands of comment letters as well as a lawsuit
against the CFTC spearheaded by two Wall Street trade groups, the
International Swaps \& Derivatives Association (ISDA) and the Securities
Industries \& Financial Markets Association (SIFMA)\footnote{ISDA \&
  SIFMA v US CFTC; Complaint, 1:11-cv-02146; December
  2\textsuperscript{nd}, 2011.}.

A world-wide debate has ensued about the role of index funds in
commodity markets. The first responses to the 2007/2008 crisis of
escalating food and energy prices took the form of policy reports, many
of which reasoned that the growth of commodity index funds came along
with an influx of largely speculative capital that was responsible for
driving commodity prices beyond their historic highs
\citep{de_schutter_food_2010_1, gilbert_speculative_2010, herman_not_2011_1, schumann_hunger-makers_2011, unctad_global_2009}.
The US and various european countries also became involved in this
issue. Early US Senate investigations drew on pricing and trading data
supplied by the CFTC as well as interviews with numerous experts
including hedge fund manager Michael Masters who linked the growing
presence of index investors in US wheat markets to the observed surges
in both futures and cash prices
\citep{masters_testimony_2008, masters_accidental_2008}. This contention
has since come to be known as the Masters' hypothesis and was largely
endorsed by the final US Senate report on the issue
\citep{USSenate_Excessive_speculation_wheat_2009}. At this stage, the
question thrust on the academic sphere was whether ``excessive
speculation'' (understood as the market activity peculiar to index type
investors) was linked to escalating energy and agricultural commodity
prices. A number of academic studies, most of which used commodity
specific causality and correlation based analysis have disproved this
contention, particularly in the case of agricultural and energy
commodities. \citet{irwin_devil_2009}, \citet{sanders_adequacy_2010},
\citet{sanders_impact_2011}, \citet{sanders_new_2011},
\citet{irwin_commodity_2013}, \citet{brunetti_commodity_2014},
\citet{hamilton_effects_2015} and \citet{bruno_financialization_2017}
study the agricultural markets. \citet{buyuksahin_speculators_2011},
\citet{tokic_speculation_2012}, \citet{fattouh_role_2013},
\citet{kilian_role_2014}, \citet{knittel_simple_2016} and
\citet{manera_modelling_2016} study the energy markets.
\citet{bohl_does_2013_2}, \citet{kim_does_2015} and
\citet{boyd_prevalence_2016} study both energy and agricultural markets
while \citet{irwin_index_2011}, \citet{irwin_financialization_2012} and
\citet{stoll_commodity_2011} study the commodity markets in general.

@@

On the one hand, futures markets have a long history of serving
commodity producers to allievate their commodity price and output risks.
\citet{keynes_treatise_1930} and \citet{hicks_value_1939} emphasise the
``normal'' behaviour of naturally short hedgers outsourcing their
commodity risk to naturally long commodity specialist speculators. This
market configuration is referred to as ``backwardation'' where futures
is expected to trade at a discount to spot, providing speculators with
incentives to take the long side. The opposite market paradigm is
referred to as ``contango''. Working
\citep{working_theory_1948, working_hedging_1953} and Cootner
\citep{cootner_returns_1960, CootnerSpeculationhedging1967} somehow
relax this ``supply-of-speculative-services''
\citep{Tilllongtermperspectivecommodity2007} approach by introducing
processors and merchants, also commodity experts, on the hedgers side
and relating the notions of backwardation and contango more directly to
the term structure via the theory of storage where the shape of the
front end of the term structure (basis) is directly related to the
market price of storage.\\
Financialization, on the other hand, refers to the entry of financial
investors into the commodity futures markets who, for the most part, are
not commodity experts. Besides, these new market participants tend to
exhibit herding like behaviour, often taking massive long-only positions
on the whole cross-section of indexed commodities in an attempt to hedge
against inflation and/or seek further diversification benefits for
portfolios with, in most cases, existing large positions across various
asset classes
\citep{brunetti_speculators_2016, boyd_prevalence_2016, cheng_convective_2014, juvenal_speculation_2015, singleton_investor_2013, tang_index_2012}.
This contrasts with the traditional approaches of Keynes and Working and
in this context the issues of co-movement and aggregate market
participants behaviour seem particularly relevant.\\
Asset pricing techniques are well suited to study co-movement. Building
on theory we develop an ad-hoc commodity based asset pricing framework
that we believe is well adapted to study these two issues simultaneously
and implement it over four periods of interest. The first, the
pre-financialization period, starts in 1997 and is naturally followed by
financialization, spanning the 2004/2007 period, and the financial
crisis and its aftermath, spanning the 2008/2013 period. The last period
spans the 2014/present period starting with the beginning of the
tapering/end of the US Quantitative Easing program (QE).\\
Using the weekly price, open interest and market position data described
above we construct market portfolios of commodity nearby futures as well
as mimicking portfolios for market, commercial hedging pressure, term
structure and liquidity risk factors in returns. While our market
portfolios are long only, equally weighted portfolios of particular
commodity sets, our other mimicking portfolios for risk factors are
constructed as combinations of two sub-portfolios (legs), one held long,
one held short, which respective constituents are pooled on a weekly
basis according to risk factor specific criterias. The weekly return for
a particular mimicking portfolio is calculated as the difference between
the average return for constituents of the long portfolio and that for
the constituents of the short portfolio. i.e, return on the mimicking
portfolio for risk factor \(i\) (\(f_{i}\)) for week \(t\):

>>>>

\documentclass[12pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[letterpaper, left = 1in, right = 1in, top = 1.5in, bottom = 1.5in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Financialization of Commodities: an Asset Pricing Perspective},
            pdfkeywords={Commodity futures; financialization; hedging pressure; financial crisis},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{natbib}
\bibliographystyle{jf}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

  \title{Financialization of Commodities: an Asset Pricing Perspective}
    % \author{Devraj Basu \\ Olivier Bauthéac}
  \author{ Devraj Basu  \\  Olivier Bauthéac  
	  \thanks{ Basu: University of Strathclyde, Glasgow, UK, \href{mailto:devraj.basu@strath.ac.uk}{\nolinkurl{devraj.basu@strath.ac.uk}}.  Bauthéac: University of Strathclyde, Glasgow, UK, \href{mailto:olivier.bautheac@strath.ac.uk}{\nolinkurl{olivier.bautheac@strath.ac.uk}}. 	  
	   \newline We thank seminar participants at IIM Ahmedabad, IIM Calcutta, SKEMA
Business School and University of Gottingen, as well as Griffith
inaugural alternative investment conference attendants for their
feedback. We also thank Joelle Miffre, Martijn Boons, John Fan, Robert
Bianchi, Chardin Wese Simen, Andrew Kaleel and particularly Olaf Korn
for their extremely useful comments and suggestions.  }
  }
    \date{}

@@

\textbf{Disclosure}

\begin{itemize}
\tightlist
\item
  Devraj Basu: I am a co-author of this paper and wish to state that I
  have no conflict of interest, financial or otherwise. I do not hold
  paid or unpaid positions at any organization whose goals relate to the
  article.\\
  \includegraphics{pics/signature - Devraj.png}\\
\end{itemize}

\textbf{Disclosure}

\begin{itemize}
\tightlist
\item
  Olivier Bauthéac: I am a co-author of this paper and wish to state
  that I have no conflict of interest, financial or otherwise. I do not
  hold paid or unpaid positions at any organization whose goals relate
  to the article.\\
  \includegraphics{pics/signature - Olivier.png}\\
\end{itemize}

@@

The last twenty years have seen major upheavals in the commodity markets
with the onset of financialization\footnote{Starting around 2004, with a
  view of commodity markets as an emerging asset class and in an effort
  to hedge against inflation and seek diversification benefits
  \citep{buyuksahin_speculators_2014, singleton_investor_2013},
  institutional investors sent forth an unprecedented flow of capital
  into commodity futures markets. This phenomenon is commonly referred
  to as the financialization of commodities
  \citep{domanski_financial_2007}.} as well as the period of the 2008
financial crisis and its aftermath. Financialization appears to have
affected a number of economic sectors, including agriculture and energy,
and its impact has been hotly debated in the policy, legislative,
regulatory, and academic spheres. A vigorous legislative debate has led
to regulatory changes accompanied by an equally intense policy and
academic debate across a number of disciplines about the perceived
effects on commodity prices of the unprecedented inflow of institutional
funds brought into commodity futures markets by new index type
investors\footnote{See section \ref{background} for a detailled
  discussion.}. In response to immediate policy concerns, the academic
debate was initially framed rather narrowly. In this paper we take a
fundamentally different empirical approach that is broader in scope. We
consider the entire cross-section of actively traded commodities on
futures markets and use a futures based asset pricing framework that
includes factors constructed using both price and non-price attributes.

Commodity futures markets have a history of being the classic physical
commodity risks shifting conduit for commodity specialists, and
financialization may have altered this long standing paradigm with the
entry of long only commodity index traders (CITs) with very different
hedging and volume demands from traditional participants. This change
could have altered the hedging ontology in these markets and our asset
pricing framework is well suited to detect these more subtle effects
driven by changes in the nature of traders. Factor model techniques are
well suited to isolating common driving factors and detecting
co-movement and building on theory we develop a bespoke commodity-based
asset pricing framework that seems well adapted to study these two
issues simultaneously.

@@

We study the performance of these models on a set of twenty-four US
traded commodities and six UK traded (LME exchange) base metals over the
1997/2017 period that we divide into four distinct sub-periods. The
first, the pre-financialization phase, spans the 1997/2003 period; the
second, the financialization phase, spans the 2004/2007 period\footnote{Starting
  point based on earlier studies
  \citep{baker_financialization_2014, christoffersen_factor_2014}.}; the
third, 2008/2013, encompasses both the financial crisis and the loose
monetary policy regimes that followed; and the fourth spans the period
starting in 2014 to present.\\
Consistent with earlier studies, we observe that mean returns for the US
traded commodities are sharply higher over the financialization period
compared to pre-financialization and this pattern extends to UK traded
metals.

The cross-sectional nature of financialization
\citep{basak_model_2016, cheng_financialization_2014} leads us to extend
the Keynesian individual hedging pressure paradigm by incorporating
aggregate CHP (\textbf{CHP}), a market wide measure of hedging pressure
first considered in \citet{hong_what_2012}. This paradigm implies that
returns for individual commodities should be high in periods where
\textbf{CHP} is low (\textbf{backwardation}) and low in periods where
\textbf{CHP} is high (\textbf{contango}). Over the first period the
cross sectional average returns during \textbf{backwardation} for both
the US and UK commodities are positive with those for the US
statistically significant, and significantly higher than during
\textbf{contango}, providing broad support for an aggregate version of
the Keynesian hedging pressure hypothesis. Over the second period this
pattern reverses, with both the US and UK commodities having significant
positive returns during \textbf{contango} with these returns being
significantly higher than during \textbf{backwardation}. This change in
pattern suggests that the onset of financialization altered the
traditional risk shifting nature of hedger's behaviour, which is implied
by the Keynesian hypothesis. The arrival of financial investors raises
the question of whether they functioned as liquidity providers or
liquidity demanders. Arguments have been made for both of these
possibilities with \citet{moskowitz_time_2012} arguing that financial
investors provide liquidity while \citet{Kangtaletwopremiums2017} argue
that they demand liquidity. Our analysis of returns in periods of low
and high aggregate CHP before and during financialization allows us to
analyse this issue from a new perspective and suggests that financial
investors may have been liquidity demanders leading hedgers to become
liquidity providers.

\citet{goldstein_commodity_2017} as well as
\citet{goldstein_speculation_2014} argue that the arrival of financial
investors has an impact on the extent of normal backwardation in
commodity futures markets. Based on their analysis we would expect our
average CHP factor, which is based on backwardation measured by hedger's
positions, as well our term structure factor, which is based on
backwardation measured by roll yield, to be best suited to measure the
impact of financialization.\\
We analyze the pricing dynamics over the different periods by first
studying the time series pricing performance of the various factors on
all the US traded commodities with the factors constructed using the
latter set of commodities. The market based factor outperforms the other
three in the first period and it has a similar level of performance in
the second period, while the CHP factor shows the greatest improvement
in performance in the second period. The performance of the market
factor is driven by agricultural commodities and metals in both the
first and second period and the dramatic improvement of the CHP factor
in the second period is driven by precious and precious-base metals. The
best performing commodities for the market factor exhibited
approximately the same level of price co-movement both before and during
financialization. In contrast, the best performing commodities for the
CHP factor exhibited very low pair-wise correlation in the first period,
around the average for US traded commodities, while during
financialization the average pairwise correlation between the top
performing CHP factor commodities is higher than that for the market
factor and double of the average for all the US traded commodities. This
increase is due to the dramatic rise in average pairwise correlation
amongst the US traded metals all of which exhibited high
R\textsuperscript{2} with respect to the CHP factor during
financialization which in turn is due to their common low level of CHP
over this period. Thus the CHP factor model is able to establish the
strengthening link between a non-price attribute and price dynamics in
the US metal sub-sector during financialization.\\
We explore this issue further by constructing the four factors with the
top eight commodities for a given factor (factor picks) in each
period\footnote{See section \ref{methods} for further details.} and
evaluating its performance on these selected commodities as well as on
all US traded commodities. The results of this exercise strongly suggest
that the hedging pressure co-movement and price co-movement link was
concentrated in the metals sub-sector.

The UK traded base metals provide a useful set of assets to test how
widespread were the changes in pricing dynamics brought about by the
onset of financialization. To that end, we examine the performance of
the four factors on the six UK metals, when constructed from each set of
factor picks. The resuslts show that here was co-movement in global
metals returns in the second period and that this co-movement could be
detected by our hedging pressure based model.

The onset of the financial crisis and the monetary policy regimes that
followed, on the other hand, also appear to have induced significant
changes in pricing dynamics. Returns on both US and UK traded
commodities fell dramatically over both the third and fourth periods
with the pattern of return for the US traded commodities across
\textbf{CHP} regimes reverting to Keynesian paradigm. The pricing
results indicate the emergence of a systematic factor across the entire
cross-section of the US commodities as well as strong evidence of
cross-market linkages.

Financialization was an issue of such policy importance that it
triggered legislative action. The debate was initially framed around
adequacy of speculation, the burning issue of the day, and the
consequent analysis focused on the more mechanical effects of
financialization. With the benefit of hindsight, it appears this
approach was perhaps too narrow and it now seems necessary to address
financialization from a broader perspective. Commodity price dynamics
appear to have altered substantially in quite different ways over the
financialization and the financial crisis periods and our commodity
futures based asset pricing approach seems able to provide new insights
into the nature of these changes: financialization was a phenomenon
endogenous to the commodity markets transmitted via the commodity
futures markets and mainly concentrated in a specific sector (metals)
while the crisis and its aftermath seems to have delivered an exogenous
shock across the entire cross-section of global commodities. The initial
view of financialization was that it consisted of speculative flows
which had the effect of driving up and creating bubble like conditions
in commodity future prices\footnote{\citet{de_schutter_food_2010_1};
  \citet{gilbert_how_2010}; \citet{gilbert_speculative_2010};
  \citet{masters_testimony_2008}; \citet{masters_accidental_2008};
  \citet{herman_not_2011_1}; \citet{schumann_hunger-makers_2011};
  \citet{singleton_investor_2013}; \citet{unctad_global_2009}.}.
However, the fundamental question of the nature of the impact of
financialization across the entire cross-section of commodity futures
markets has not yet been completely answered. We provide an empirical
complement to a new stream of theoretical studies that try to model the
impact of financialization on various aspects of commodity futures
markets\footnote{\citet{etula_broker-dealer_2013},
  \citet{acharya_limits_2013}, \citet{cheng_convective_2014},
  \citet{leclercq_equilibrium_2014}, \citet{sockin_informational_2015},
  \citet{goldstein_speculation_2013},
  \citet{EkelandSpeculationcommodityfutures2016},
  \citet{goldstein_commodity_2017}.} and demonstrate that the effects of
financialization extend beyond the mechanical effects induced by
indexation as outlined in \citet{basak_model_2016} as well as
\citet{tang_index_2012}.

@@

The period from 1998 to 2008 saw an influx of investment into commodity
index linked products most of which made its way into the commodity
futures market. These commodity indexes whose goal was to track the
broad movement of commodity prices could be accessed by means of
exchange traded funds and exchange traded notes. At least \$100 billion
of new investment flowed into these products over the 1998-2008 period
with this financialization phenomenon leading to substantial legislative
changes. Concerns over the consequences of financialization were behind
Rule 76 FR 4752 issued by the US Commodity Futures Trading Commission
(CFTC) on January 26\textsuperscript{th}, 2011. This provision emanates
from the Dodd-Frank Wall Street and Consumer Protection Act of 2010
(Title VII, Section 737) that mandates the CFTC to use position limits
to restrict the flow of speculative capital into a number of commodity
markets. The Rule was approved in a close 3-2 vote and the ensuing
rulemaking process was extremely contentious with several commissioners
(Michael Dunn and Scott D. O'Malia in particular) expressing
reservations about the lack of supporting evidence and the Rule also
triggering thousands of comment letters as well as a lawsuit against the
CFTC spearheaded by two Wall Street trade groups, the International
Swaps \& Derivatives Association (ISDA) and the Securities Industries \&
Financial Markets Association (SIFMA)\footnote{ISDA \& SIFMA v US CFTC;
  Complaint, 1:11-cv-02146; December 2\textsuperscript{nd}, 2011.}.

A world-wide debate has ensued about the role of index funds in
commodity markets. The first responses to the 2007/2008 crisis of
escalating food and energy prices took the form of policy reports, many
of which reasoned that the growth of commodity index funds came along
with an influx of largely speculative capital that was responsible for
driving commodity prices beyond their historic highs
\citep{de_schutter_food_2010_1, gilbert_speculative_2010, herman_not_2011_1, schumann_hunger-makers_2011, unctad_global_2009}.
The US and various european countries also became involved in this
issue. Early US Senate investigations drew on pricing and trading data
supplied by the CFTC as well as interviews with numerous experts
including hedge fund manager Michael Masters who linked the growing
presence of index investors in US wheat markets to the observed surges
in both futures and cash prices
\citep{masters_testimony_2008, masters_accidental_2008}). This
contention has since come to be known as the Masters' hypothesis and was
largely endorsed by the final US Senate report on the issue
\citep{USSenate_Excessive_speculation_wheat_2009}. At this stage, the
question thrust on the academic sphere was whether ``excessive
speculation'' (understood as the market activity peculiar to index type
investors) was linked to escalating energy and agricultural commodity
prices. A number of academic studies, most of which used commodity
specific causality and correlation based analysis have disproved this
contention, particularly in the case of agricultural and energy
commodities. \citet{irwin_devil_2009}, \citet{sanders_adequacy_2010},
\citet{sanders_impact_2011}, \citet{sanders_new_2011},
\citet{irwin_commodity_2013}, \citet{brunetti_commodity_2014},
\citet{hamilton_effects_2015} and \citet{bruno_financialization_2017}
study the agricultural markets. \citet{buyuksahin_speculators_2011},
\citet{tokic_speculation_2012}, \citet{fattouh_role_2013},
\citet{kilian_role_2014}, \citet{knittel_simple_2016} and
\citet{manera_modelling_2016} study the energy markets.
\citet{bohl_does_2013_2}, \citet{kim_does_2015} and
\citet{boyd_prevalence_2016} study both energy and agricultural markets
while \citet{irwin_index_2011}, \citet{irwin_financialization_2012} and
\citet{stoll_commodity_2011} study the commodity markets in general.

@@

On the one hand, futures markets have a long history of serving
commodity producers to allievate their commodity price and output risks.
\citet{keynes_treatise_1930} and \citet{hicks_value_1939} emphasise the
``normal'' behaviour of naturally short hedgers outsourcing their
commodity risk to naturally long commodity specialist speculators. This
market configuration is referred to as ``backwardation'' where futures
is expected to trade at a discount to spot, providing speculators with
incentives to take the long side. The opposite market paradigm is
referred to as ``contango''. Working
\citep{working_theory_1948, working_hedging_1953} and Cootner
\citep{cootner_returns_1960, CootnerSpeculationhedging1967} somehow
relax this ``supply-of-speculative-services''
\citep{Tilllongtermperspectivecommodity2007} approach by introducing
processors and merchants, also commodity experts, on the hedgers side
and relating the notions of backwardation and contango more directly to
the term structure via the theory of storage where the shape of the
front end of the term structure (basis) is directly related to the
market price of storage.\\
Financialization, on the other hand, refers to the entry of financial
investors into the commodity futures markets who, for the most part, are
not commodity experts. Besides, these new market participants tend to
exhibit herding like behaviour, often taking massive long-only positions
on the whole cross-section of indexed commodities in an attempt to hedge
against inflation and/or seek further diversification benefits for
portfolios with, in most cases, existing large positions across various
asset classes
\citep{brunetti_speculators_2016, boyd_prevalence_2016, cheng_convective_2014, juvenal_speculation_2015, singleton_investor_2013, tang_index_2012}.
This contrasts with the traditional approaches of Keynes and Working and
in this context the issues of co-movement and aggregate market
participants behaviour seem particularly relevant.\\
Asset pricing techniques are well suited to study co-movement. Building
on theory we develop an ad-hoc commodity based asset pricing framework
that we believe is well adapted to study these two issues simultaneously
and implement it over four periods of interest. The first, the
pre-financialization period, starts in 1997 and is naturally followed by
financialization, spanning the 2004/2007 period, and the financial
crisis and its aftermath, spanning the 2008/2013 period. The last period
spans the 2014/present period sarting with the beginning of the
tapering/end of the US Quantitative Easing program (QE).\\
Using the weekly price, open interest and market position data described
above we construct market portfolios of commodity nearby futures as well
as mimicking portfolios for market, commercial hedging pressure, term
structure and liquidity risk factors in returns. While our market
portfolios are long only, equally weighted portfolios of particular
commodity sets, our other mimicking portfolios for risk factors are
constructed as combinations of two sub-portfolios (legs), one held long,
one held short, which respective constituents are pooled on a weekly
basis according to risk factor specific criterias. The weekly return for
a particular mimicking portfolio is calculated as the difference between
the average return for constituents of the long portfolio and that for
the constituents of the short portfolio. i.e, return on the mimicking
portfolio for risk factor \(i\) (\(f_{i}\)) for week \(t\):
